Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-09-13 Writing/NatureWater/2ndRoundRevision/replies/reply_to_reviewers_R2.tex

\textcolor{blue}{Thank you for catching this. We have fixed all instances of this typo.}

\item Figure 1 caption – I found the term ``county-year water deficit values'' confusing. I recommend use of something like ``county-level average water deficit values''. Might also consider modifying line 290.

\textcolor{blue}{We agree. Instead of county-level "average" water deficit, in our revised manuscript we now say county-level annual water deficit because they are not averaged values.}

\textcolor{blue}{We also changed "evident in county-years with large water deficits" in line 290 in the previous manuscript to "evident in years with large water deficits," which is in line 275 in the current manuscript.}

@@

\item New comment 2 – Figure 2 results below 200mm: Lines 99 states that quantiles 3 has ``statistically lower yields''. Please clarify this. From visual inspection of Figure 2, the quantile 3 yields do not seem to be significantly below the quantile 1 values (at least for corn). Could you please clarify this? If the result is not statistically significant, does it make sense to discuss it in lines 98-111? Since the paper otherwise applies statistical significance as a criterion for meaningful results, I would recommend being consistent.

    \textcolor{blue}{Thank you for noticing this. You are correct that this difference is no longer statistically significant. Following your suggestion, we have removed this paragraph from the results to ensure consistency in what is discussed as significant in the context of our analysis.}

\item New comment 3 - Figure 4 category changes: Why were the aquifer thickness categories changed in Figure 4? Please also clarify how function estimates for point values of aquifer thickness (rather than value ranges) are derived from the data and explain what motivates using different categories in Figures 1 and 4.

    \textcolor{blue}{Thank you for this helpful point, which highlgihts a need to clarify some details about how aquifer thickness enters different parts of our analysis. While the irrigated yield regression (equation 1, Figure 1) uses aquifer thickness category, the irrigated share regression (Figure 4) uses aquifer thickness as it is as a continuous variable. In our average productivity analysis (Figure 4) which combines these two regressions, aquifer thickness is specified as a specific point value for the component related to irrigated area share and as a categorical value for component related to irrigated per-area yields. This step has been the same throughout all versions of our manuscript, and is not a change that has been introduced in current or prior revisions. We have added the following sentence on line 341 of our methods to clarify this point:}

@@

    \textcolor{blue}{In the previous version of our average productivity analysis, we assess the impacts of four aquifer thickness values: 10, 50, 90, and 130 meters in the initial version of our manuscript, with equal spacing chosen to illustrate non-linearity in impacts of aquifer thickness. In our last round of revisions, we changed these values to instead be 10, 40, 70, and 100m. These values ensure that we have a range of values spanning different aquifer thickness categories. Please note that key qualitative message and findings is identical whichever set of values we use. As we mentioned earlier, we made this change to illustrate that significant non-linear changes in total productivity can happen within a narrower range of evenly spaced aquifer thickness compared to the previous sets of aquifer thickness values used in this figure. This change also provide results that are more realistically centered on the range of aquifer thickness values observed within our study area and period.}

\item New comment 4 – causality and fixed effects: The manuscript has been strengthened by the expansion of the section explaining the reasons for assuming declining aquifer thickness causally reduces the ability to use irrigation to maintain crop yields (lines 40-57). Please clarify the following:

  \begin{itemize}
  \item (4.a) Please clarify for readers if other major causal pathways might be relevant. For example: Are there relevant regulations that limit abstractions in counties where aquifers are too depleted? If so, the causality would be similar, but the implications somewhat different. If not, then stating this would be helpful so readers can rule out potentially strong alternative causal relationships.

  \textcolor{blue}{Regulatory pumping restrictions have limited impacts on water use decisions in our study area, and therefore are unlikely to impact the ability to use groundwater as a buffer against drought. In contrast, the impacts of aquifer depletion on farmers' ability to meet crop water needs through changes in well yields are linked directly to hydrologic and agronomic theory, and have demonstrated extensively through prior work cited in our manuscript. While we note that restrictions in pumping are imposed in some parts of the High Plains Aquifer, there are several reasons why we can confidently state these would not be a confounding source of causality in our analysis. Please see the list below. We added the following paragraphs to a new Appendix A in this round of manuscript, as we feel these are important points to include but are constrained by word limits in the main manuscript.}

@@

  \item (4.b) Relatedly, the responses to my previous comments 1.a and 7.b indicated that county-level fixed effects were not included in any regressions (only state-level). The response to reviewer \#2 and the Methods section suggest that county-level fixed effects were used in Equation 1. Please clarify the discrepancy and make sure this is clear in the paper.

  \textcolor{blue}{In our initial submission, we noted in our methods that county fixed effects $\alpha_i$ were included in equation 1 but not in equation 2. The reason for this difference is outlined in Appendix C, which provides a detailed discussion of why including county fixed effects for our irrigated area share regression is problematic.}

  \item (4.c) The following conclusion in Appendix C, lines 410-414 does not seem fully convincing in its present form: The fact that state-level fixed effects are not influential does not immediately imply that county-level fixed effects are ``unlikely to cause significant bias''. There could e.g. easily be relevant county-level effects that average out at the state level.

  \textcolor{blue}{We recognize your point that this may be potentially too strong of a claim. To more carefully summarise the implications of our choice of fixed effects, we have modified the text on lines 448-453 for our revision of the statement.}

@@

  \textcolor{blue}{"This indicates that, even though there may be unobservables that are spatially correlated with aquifer thickness beyond soil characteristics, their correlation may not cause significant bias to our estimation and findings. We note that more detailed assessment of the impacts of including county-fixed effects is beyond the scope of our available datasets, and we defer this matter to future studies where greater temporal variations in aquifer thickness levels may be available to allow incorporation of county-level fixed effects."}

  \textcolor{blue}{We also removed the sentence below that was on lines 345 - 346 in the previous version of our manuscript to reflect that we can only hypothesize about the likely differences between using state and county fixed effects:}

@@

\subsection{MINOR COMMENTS}

\begin{itemize}

\item (M1) The response letter indicates that ``all [minor] points have been addressed''. Most minor comments were addressed correctly. The following minor comments from the previous round, however, have not yet been addressed:

\textcolor{blue}{We apologize for this mistake. This time, all the minor comments are addressed.}

@@

\textcolor{blue}{``$\mu_t$'' represents year fixed effects. However, as we now use state-year FE represented by $\theta_{s,t}$, the ``$\mu_t$' should have been removed. Equation 2 no longer includes $\mu_t$.}

\item (M2) Some additional minor points have arisen:

\item (M2.a) Line 315: ``Y'' is capitalized here, but not in equation (1)

\textcolor{blue}{We changed $y_{i,t}$ in equation (1) to $Y_{i,t}$.}

### 2023-09-13 Writing/NatureWater/2ndRoundRevision/manuscript.tex
second round submission (finalized)

where county and year are indicated by subscripts \(i\) and \(t\), respectively, \(y\) is crop yield, and \(D\) is seasonal water deficit. \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: rainfed production, and county-years with aquifer thickness falling in the \([0\%-33.3\%)\), \([33.3\%-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of aquifer thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner, with \(\sigma_j\) representing the category specific intercept, \(\phi_{j,k}\) \(k\)th spline basis function, and \(\beta_k\) is its coefficient (see Appendix \ref{spline-basis} for further explanations on how spline-based regression works). $K$ is the total number of spline basis functions. $K = 3$ was chosen for this regression as it minimizes Bayesian Information Criterion (BIC). To control for time-invariant heterogeneity across counties and yearly shocks, county fixed effects (\(\alpha_i\)) and year fixed effects (\(\lambda_t\)) are also included and standard error estimation is clustered by county.

@@

Some parts of the HPA region face regulatory pumping restrictions in response to declines in aquifer thickness. Resulting groundwater use reductions can in turn negatively affect crop yield. This causal pathway of aquifer thickness affecting crop yield is distinct from the pathways via well yield discussed in the \ref{main} section. However, it is unlikely that our regression analysis capture this effect.

The majority of groundwater withdrawal regulations that do exist in the HPA limit pumping to a specified volume over a multiyear period, and farmers have also been allowed to bank and carryover significant volumes of water deficits from historic periods when allocation levels were higher \citep{rimvsaite2023groundwater}. This enables farmers to increase water withdrawals substantially in drought years when demand is higher, and prior research has shown that farmers' water use decisions are not constrained by such regulations even in extreme drought years such as 2012 \citep{foster2019assessing}.

Where more restrictive groundwater withdrawal limits have been introduced, for example in some Groundwater Management Districts in Kansas \citep{marston2022, rimvsaite2023groundwater}, these have universally been implemented in the last 5-10 years and therefore have no or only minimal overlap with our study period which extends up to 2016. The earliest of these groundwater managements in Kansas interventions was the Sheridan 6 Local Enhanced Management Area (LEMA) in NW Kansas, which introduced restrictions to farmer groundwater pumping of 11 inches per year on average per 5-year period starting in 2013. While this overlaps with the final 4 years of our study period, the LEMA area covers only parts of two counties and hence represents a very small proportion of our overall study sample while the years 2013-2016 were all characterized by average or above average growing season rainfall (i.e. not drought years). Moreover, evidence shows that the effect of restrictions introduced by the LEMA has primarily been a reduction in per-area water use, with irrigated crop yields largely unaffected and only small reductions in irrigated areas compared to pre-policy conditions \citep{drysdale2018adaptation, deines2021combining}. These policies therefore would not be expected to confound our results on the impacts of aquifer thickness.

Areas of the southern HPA, in particular in northern Texas, where the lowest aquifer thickness values are observed have few if any regulatory restrictions on groundwater extraction due to laws that typically provide farmers with absolute rights to use groundwater under their land coupled with low levels of metering that constrain potential for implementing extensive and binding pumping restrictions \citep{wheeler2016lessons, closas2018chronicle}.

>>>>

where county and year are indicated by subscripts \(i\) and \(t\), respectively, \(y\) is crop yield, and \(D\) is seasonal water deficit. \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: rainfed production, and county-years with aquifer thickness falling in the \([0\%-33.3\%)\), \([33.3\%-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of aquifer thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner, with \(\sigma_j\) representing the category-specific intercept, \(\phi_{j,k}\) \(k\)th spline basis function, and \(\beta_k\) is its coefficient (see Appendix \ref{spline-basis} for further explanations on how spline-based regression works). $K$ is the total number of spline basis functions. $K = 3$ was chosen for this regression as it minimizes Bayesian Information Criterion (BIC). To control for time-invariant heterogeneity across counties and yearly shocks, county fixed effects (\(\alpha_i\)) and year fixed effects (\(\lambda_t\)) are also included and standard error estimation is clustered by county.

@@

Our analysis evaluates the links between aquifer thickness and drought resilience, proposing a causal link between declining aquifer thickness and irrigated production during droughts. In the first section of our manuscript, we propose that this connection is driven by impacts of aquifer thickness on well yields and, hence, farmers ability to use groundwater to fully buffer crops against water deficits during drought.

It is important to assess the extent to which this causal relationship may be affected or confounded by other factors. One key alternative hypothesis is that regulations limiting groundwater withdrawals could have been introduced as a response to declining aquifer thickness, with resulting reductions in irrigated areas and irrigated crop yields driven by regulatory rather than hydrologic (i.e. well yield) constraints. Below, we discuss the reasons that regulatory constraints are highly unlikely to be the cause of the relationships between aquifer thickness and drought impacts observed in our analysis for the HPA.

First, it is important the majority of groundwater withdrawal regulations that do exist in the HPA limit pumping to a specified volume over a multiyear period. It is also common for farmers to have been allowed to bank and carryover significant volumes of water deficits from historic periods when allocation levels were higher \citep{rimvsaite2023groundwater}, further increasing their regulatory quotas above defined caps. Together, these factors mean that farmers in parts of the HPA where groundwater withdrawals are regulated or restricted are still able to increase water withdrawals substantially in drought years when demand is higher. Prior research has shown that as a result farmers water use decisions in HPA are not constrained by such regulations even in extreme drought years such as 2012 \citep{foster2019assessing}.

Where more restrictive groundwater withdrawal limits have been introduced, for example in some Groundwater Management Districts in Kansas \citep{marston2022, rimvsaite2023groundwater}, these have universally been implemented in the last 5-10 years and therefore have no or only minimal overlap with our study period which extends up to only 2016. The earliest of these groundwater managements in Kansas interventions was the Sheridan 6 Local Enhanced Management Area (LEMA) in Northwest Kansas, which introduced restrictions to farmer groundwater pumping of 11 inches per year on average per 5-year period starting in 2013. While this overlaps with the final 4 years of our study period, the Sheridan 6 LEMA area covers only parts of two counties in Kansas and hence represents a very small proportion of our overall study sample. Additionally, the years 2013-2016 were all characterized by average or above average growing season rainfall (i.e. not drought years). Furthermore, evidence shows that the effect of restrictions introduced by the LEMA has primarily been a reduction in per-area water use, with irrigated crop yields largely unaffected and only small reductions in irrigated areas compared to pre-policy conditions \citep{drysdale2018adaptation, deines2021combining}. These policies therefore would not be expected to confound our results on the impacts of aquifer thickness.

Finally, we also highlight that areas of the southern HPA, in particular in northern Texas, where the lowest aquifer thickness values are observed have few if any regulatory restrictions on groundwater extraction. This is due to laws that typically provide farmers with absolute rights to use groundwater under their land, coupled with low levels of metering. Research has shown that these conditions have historically limited or constrained potential for implementing extensive and binding pumping restrictions on irrigators in these areas \citep{wheeler2016lessons, closas2018chronicle}, again indicating that regulations are not the binding driver of water use constraints in these areas. In contrast, as discussed in the \ref{main} section of our manuscript, there is extensive evidence highlighting the reductions in well yields that have occurred in these parts of the HPA along with anecdotal, theoretical and simulation based evidence for why these restrictions would impose meaningful restrictions on farmers ability to adequately meet crop water needs during droughts.
